---
title: "01 - Building the RCTD Reference from the Pelka et al. CRC Atlas"
output: github_document
---

## What this notebook does, and why it's separate

Cell-type deconvolution (Section 8 of the main analysis) needs an
**annotated single-cell reference**: real cells of known type, so RCTD
can learn what each cell type's expression profile looks like and then
ask "what mixture of these profiles best explains each spatial bin?"

That reference is a substantial dataset in its own right (371,223 cells),
built once and reused, so it gets its own notebook rather than being
buried inside the main pipeline -- exactly the same reason the original
workshop material this project is modeled on treated reference-building
as a separate pre-step rather than inline code.

We use [Pelka et al. 2021, *Cell*](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE178341) --
human colorectal tumor **and** normal-adjacent cells, annotated by cell
type, from the same tissue type as our Visium HD samples. Matching
reference tissue to spatial tissue matters: deconvolving human colon
spots against, say, a brain atlas would force RCTD to explain every spot
with cell types that were never actually there.

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(Seurat)
  library(spacexr)
  library(tidyverse)
  library(data.table)
  library(hdf5r)
})

set.seed(123)

ref_dir <- "../data/scref"
out_dir <- "../output/rdata"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

h5_path       <- file.path(ref_dir, "crc10x_full_c295v4_submit.h5")
cluster_path  <- file.path(ref_dir, "crc10x_full_c295v4_submit_cluster.csv.gz")
metatable_path <- file.path(ref_dir, "crc10x_full_c295v4_submit_metatables.csv.gz")

stopifnot(
  "Run download_data.sh first -- reference files not found" =
    file.exists(h5_path, cluster_path, metatable_path)
)
```

## Step 1: Load the expression matrix

GEO deposits vary in internal H5 structure -- some are plain CellRanger-
style matrices (`Read10X_h5` handles these directly), others are
AnnData/`SummarizedExperiment`-style objects with different internal
group names. We try the standard loader first and fall back to
inspecting the file's structure by hand if it doesn't fit, rather than
assuming and failing silently.

```{r load-h5}
counts_ref <- tryCatch(
  Read10X_h5(h5_path),
  error = function(e) {
    message("Read10X_h5() didn't work directly (", conditionMessage(e), ").\n",
            "Inspecting the file's internal structure instead:")
    f <- H5File$new(h5_path, mode = "r")
    print(f$ls(recursive = TRUE))
    f$close_all()
    stop(
      "See the group listing above, adapt the loading code to match it ",
      "(common alternative: matrix/genes/barcodes stored under different ",
      "group names -- build the dgCMatrix by hand from those groups), ",
      "then re-run this chunk."
    )
  }
)

cat("Reference matrix:", nrow(counts_ref), "genes x", ncol(counts_ref), "cells\n")
```

## Step 2: Load and inspect the cell-type metadata

We don't hard-code a column name here -- we print what's actually in the
file and pick the cell-type column deliberately, which is the right
habit for any dataset you didn't generate yourself.

```{r load-metadata}
cluster_tbl  <- fread(cluster_path)
metatable    <- fread(metatable_path)

cat("cluster file columns:", paste(colnames(cluster_tbl), collapse = ", "), "\n")
cat("metatable columns:   ", paste(colnames(metatable), collapse = ", "), "\n")

head(cluster_tbl)
head(metatable)
```

```{r pick-celltype-column}
# Real-world GEO quirk, confirmed against this release: the two files
# split information differently than their names suggest. `metatable`
# carries per-cell *clinical* fields (tumor stage, MMR status, tissue
# site, patient ID...) but no cell-type label at all -- the cell-type
# calls (clTopLevel = broad lineage, cl295v11SubFull = fine subtype)
# are in `cluster_tbl` instead. And cluster_tbl's first column, whatever
# fread happens to name it, actually holds the cell barcode (values look
# like "C103_T_1_1_0_c1_v2_id-AAACCTGCATGCTAGT", matching
# metatable$cellID) -- not what its header implies. So: read that first
# column by position, not by trusting its name, and take cell type from
# cluster_tbl rather than metatable.
cluster_barcode <- cluster_tbl[[1]]

celltype_candidates <- c("clTopLevel", "cl295v11SubFull", "cl295v11SubShort", "clMidwayPr", "cellType", "cell_type", "CellType")
celltype_col <- intersect(celltype_candidates, colnames(cluster_tbl))[1]

stopifnot(
  "None of the expected cell-type column names were found in cluster_tbl -- open \
   the cluster file printout above and set `celltype_col` to the right column \
   name by hand." = !is.na(celltype_col)
)
cat("Using '", celltype_col, "' (from the cluster file) as the cell-type label.\n", sep = "")

meta <- tibble(cell_id = cluster_barcode, cell_type = cluster_tbl[[celltype_col]]) %>%
  filter(cell_id %in% colnames(counts_ref)) %>%
  distinct(cell_id, .keep_all = TRUE)

cat("Matched", nrow(meta), "of", ncol(counts_ref), "reference cells to a cell-type label.\n")
cat(sum(meta$cell_id %in% metatable$cellID), "of those also have matching clinical metadata in metatable.\n")
```

## Step 3: Filter rare cell types

RCTD needs a reasonable number of example cells per type to learn a
stable expression profile -- the same `>= 10 cells` bar used for every
reference in the main analysis.

```{r filter-rare-types}
counts_ref <- counts_ref[, meta$cell_id]
cell_types_full <- setNames(factor(meta$cell_type), meta$cell_id)

ct_counts  <- table(cell_types_full)
keep_types <- names(ct_counts[ct_counts >= 10])
cat("Dropping", sum(ct_counts < 10), "cell types with <10 cells (of",
    length(ct_counts), "total types)\n")

keep_cells   <- names(cell_types_full)[cell_types_full %in% keep_types]
counts_ref   <- counts_ref[, keep_cells]
cell_types   <- droplevels(cell_types_full[keep_cells])

cat("Final reference:", ncol(counts_ref), "cells across", nlevels(cell_types), "cell types\n")
print(sort(table(cell_types), decreasing = TRUE))
```

## Step 4: Build the RCTD `Reference` object

```{r build-reference}
nUMI_ref <- colSums(counts_ref)
rctd_ref <- Reference(counts_ref, cell_types, nUMI_ref)
```

## Step 5: Save for reuse

```{r save}
if (requireNamespace("qs2", quietly = TRUE)) {
  qs2::qs_save(rctd_ref, file.path(out_dir, "crc_rctd_reference.qs2"))
} else {
  saveRDS(rctd_ref, file.path(out_dir, "crc_rctd_reference.rds"))
}
cat("Saved. This is the object 02_visium_hd_analysis.Rmd loads for RCTD.\n")
```
